Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/splitting.tex
4.3 all the way to next/prev

\paragraph {Injections.}
Now consider splitting the non-terminal, $\inl {e}$, where $\diaone
{e} \xi v$ and $\rtab \xi v \vsplito \mval i q$. Therefore:
\[
	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
\]
Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
splitting must produce a $c'$ and $l.r'$ satisfying,
\[
	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
\]
Our algorithm generates a first-pass term $c'$ that produces the same
boundary value ($b$) as the split of the subterm $e$:
\[
c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}
\]
Splitting also generates the same resumer as the split of the
subterm ($l.r' = l.r$)

\paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
where $\diaone {e_k} {\xi_k} {v_k}$ and 
$\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$ for $k = 1,2$.  
Inductively:
\[
	\splitone {e_k} A {c_k} {l_k.r_k} \text{ and } \reduce {c_k} {\tup{i_k,b_k}} \text{ and } [b_k/l_k]r_k \equiv q_k
\]
Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}}$ and $\tup{v_1,v_2} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
splitting generates:
\[
	\reduce {c} {\tup{\tup{i_1,i_2},b}} \text{ and } [b/l]r \equiv \tup{q_1,q_2}
\]
The resulting boundary value is the tuple of the boundary values of
the subterms, $b = \tup{b_1,b_2}$.  The value is 
constructed using,
\[
c = \letin{\tup{y_1,z_1}}{c_1}{\letin{\tup{y_2,z_2}}{c_2}{\tup{\tup{y_1, y_2},\tup{z_1, z_2}}}}
\]
And deconstructed with a tuple pattern, $l.r = \tup{l_1,l_2}.\tup{r_1,r_2}$,

\paragraph {Case}
Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
Here we consider only the case where
$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
(The \texttt{inr} case is analogous.)

Inductively,
\[
	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
\]
Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$,
and so the entire case evaluates like:
$\diaone {\caseof {e_1} {x_2.e_2} {x_3.e_3}} {\xi_1,\xi_2} {v_2}$.

Inductively,
\[
	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2} {\tup{i_2,b_2}} 
\]
\[
\text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
\]
(Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
When we substitute some $v_1$ for $x_2$ in $e_2$, we must substitute the masks
of $v_1$ into $c_2$ and $l_2.r_2$.)

The invariant, applied to the entire \texttt{case}, requires some $c'$ and
$l'.r'$ such that 
\[
  \reduce {c'} {\tup{i_2,b'}} \text{ and }
  [b'/l']r' \equiv q_2
\]
We choose
\[
c' =
{\left(
\talllet{\tup{y_1,z_1}}{c_1}{
\tallcase{y_1}
{x_2.\letin{\tup{y_2,z_2}}{c_2}{\tup{y_2,\tup{z_1,\inl{z_2}}}}}
{x_3.\cdots\ttrpar}
}\right)}
\]
\[
l'.r' =
\tup{l_1,l_b}.\letin{z}{r_1}{\caseof{l_b}{l_2.[z/x_2]{r_2}}{l_3.\cdots}}
\]
If you work it out you will see that $c'$ reduces to
$\tup{i_2,\tup{b_1,\inl{b_2}}}$, meaning that 
$b' = \tup{b_1,\inl{b_2}}$, and
$[b'/l']r'$ essentially reduces to
$[[b_1/l_1]r_1/x_2][b_2/l_2]r_2$, which is exactly the 
$[q_1/x_2][b_2/l_2]r_2$ we expected.

In other words, $c'$ cases on $i_1$ to run the first branch of first-stage
code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
{b_2}}$. The resumer of the predicate uses $b_1$, and $\inl{b_2}$ signals to run
the first branch's resumer. The second branches of $c'$ and $l'.r'$ are
similarly constrained by the $\texttt{inr}$ case.

@@

\paragraph {Function} 
Function introduction has a $\tup{}$ boundary value,
since functions are already fully reduced in our semantics.
However, since the body of a function may itself be multistage, splitting must continue into it.
The $i$ is a new function formed from the first-stage part of the original body.
The resumer is a new function formed out of the second-stage part of the original body.
It is the responsibility of the application site to save the precomputation of the function body
and pass it to the resumer version of the function.

@@

Because world \bbtwo\ terms in \lang\ reduce to monostage residuals (as opposed to partial values),
term splitting at world \bbtwo\ is simpler and more uniform than the version at \bbonem. 
The algorithm is specified by the $\splittwosym$ relation in \cref{fig:termSplit}.

Consider any (except \texttt{prev}) 
$n$-ary term $e = \scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$.
First note that for any $\diatwo e q$, $q$ must have the form $\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$,
even under binders and for nullary terms.
For each $k \in \{1,\ldots,n\}$, the invariant for stage \bbtwo\ splitting of $e_k$ is,
\[
	\splittwo {e_k} A {p_k} {l_k} {r_k} \text{ and } \reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
\]
To match the invariant for a split $\splittwo {\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}} A p l r$, we require,
\[
	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
\]
Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
$l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.

Again, this holds even for nullary terms and under binders.
In particular, \texttt{case} splits according to the rule:
\[
\infersplittwo	{	\begin{array} {l}
						\caseof{e_1}{x_2.e_2}{x_3.e_3} \splittwosym \\
						\hspace{5em} \pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}		
					\end{array}}	
				{\sub 1 ? & \sub 2 ? & \sub 3 ?} 								
\]
Note how the precomputations are pulled out from underneath the binder into a tuple that is always executed in the first stage,
regardless of what branch is eventually executed in the second stage.

\paragraph{Next and Prev}
Since \texttt{next} and \texttt{prev} do nothing more than convert between 
second-stage terms at \bbtwo\ and second-stage terms encapsulated within $\fut$s at \bbonem,
their splitting rules are relatively simple.
Given an $\splittwo e ? p l r$ and $\diatwo e q$, the splitting invariant says,
\[
	\reduce p b \text{ and } [b/l]r \equiv q
\]
The splitting invariant for $\splitone {\next {e'}} ? {c'} {l'.r'}$,
where $\diaone {\next {e'}} {{\hat y} \mapsto q} {\next {\hat y}}$
and $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$
then requires
\[
	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
\]
which can be fulfilled using $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.

>>>>

\paragraph {Injections.}
Now consider splitting the non-terminal, $\inl {e}$, where $\diaone
{e} \xi v$ and $\rtab \xi v \vsplito \mval i q$. Therefore:
\[
	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
\]
Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
splitting must produce a $c'$ and $l.r'$ satisfying,
\[
	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
\]
Our algorithm generates a first-pass term $c'$ that produces the same
boundary value as splitting $e$:
\[
c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}
\]
Splitting the injection also generates the same resumer as splitting
the subterm $e$: ($l.r' = l.r$).

\paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
where $\diaone {e_k} {\xi_k} {v_k}$ and 
$\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$ for $k = 1,2$.  
Inductively:
\[
	\splitone {e_k} A {c_k} {l_k.r_k} \text{ and } \reduce {c_k} {\tup{i_k,b_k}} \text{ and } [b_k/l_k]r_k \equiv q_k
\]
Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}}$ and $\tup{v_1,v_2} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
splitting generates:
\[
	\reduce {c} {\tup{\tup{i_1,i_2},b}} \text{ and } [b/l]r \equiv \tup{q_1,q_2}
\]
The resulting boundary value is the tuple of the boundary values of
the subterms, $b = \tup{b_1,b_2}$.  The value is 
constructed using,
\[
c = \letin{\tup{y_1,z_1}}{c_1}{\letin{\tup{y_2,z_2}}{c_2}{\tup{\tup{y_1, y_2},\tup{z_1, z_2}}}}
\]
And deconstructed in the resumer with the tuple pattern,
\[
l.r = \tup{l_1,l_2}.\tup{r_1,r_2}
\]
\paragraph {Case.}
Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
Here we consider only the case where
$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and \\ $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
(The \texttt{inr} case is analogous.)

\noindent
Inductively,
\[
	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
\]
Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
Therefore:
\[
	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2} {\tup{i_2,b_2}} 
\]
\[
\text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
\]
(Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
When we substitute some $v_1$ for $x_2$ in $e_2$, we must substitute the masks
of $v_1$ into $c_2$ and $l_2.r_2$.)

The entire case evaluates as
$\diaone {\caseof {e_1} {x_2.e_2} {x_3.e_3}} {\xi_1,\xi_2} {v_2}$, so splitting should produce 
some $c'$ and
$l'.r'$ such that 
\[
  \reduce {c'} {\tup{i_2,b'}} \text{ and }
  [b'/l']r' \equiv q_2
\]
We choose
\[
c' =
{\left(
\talllet{\tup{y_1,z_1}}{c_1}{
\tallcase{y_1}
{x_2.\letin{\tup{y_2,z_2}}{c_2}{\tup{y_2,\tup{z_1,\inl{z_2}}}}}
{x_3.\cdots\ttrpar}
}\right)}
\]
\[
l'.r' =
\tup{l_1,l_b}.\letin{z}{r_1}{\caseof{l_b}{l_2.[z/x_2]{r_2}}{l_3.\cdots}}
\]
The term $c'$ reduces to
$\tup{i_2,\tup{b_1,\inl{b_2}}}$, so 
$b' = \tup{b_1,\inl{b_2}}$. Therefore
$[b'/l']r'$ reduces to
$[[b_1/l_1]r_1/x_2][b_2/l_2]r_2$, exactly the 
$[q_1/x_2][b_2/l_2]r_2$ required.

More intuitively, $c'$ cases on $i_1$ to run the first branch of first-stage
code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
{b_2}}$. The resumer of the predicate uses $b_1$, and $\inl{b_2}$ signals to run
the first branch's resumer. The second branches of $c'$ and $l'.r'$ are
similarly constrained by the $\texttt{inr}$ case.

@@

\paragraph {Function.} 
Function introduction has a $\tup{}$ boundary value, since functions
are already fully reduced in our semantics.  However, since the body
of a function may itself be multistage, splitting must continue into
it.  The first-pass term generate by split produces an $i$ that is a
new function formed from the first-stage part of the original body.
The resumer is a new function formed out of the second-stage part of
the original body.  It is the responsibility of the application site
to save the precomputation of the function body and pass it to the
resumer version of the function.

@@

Because world \bbtwo\ terms in \lang\ reduce to monostage residuals
(as opposed to partial values), term splitting at world
\bbtwo\ (specified by the $\splittwosym$ relation in
\cref{fig:termSplit}) is simpler than that of \bbonem.

Consider any $n$-ary \bbtwo-term $e = \scriptCapp {e_1 \ttsemi \ldots
  \ttsemi e_n}$ (except \texttt{prev}, which is discussed shortly).
Note that for any $\diatwo e q$, the residual $q$ must have the form
$\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even under binders and
for nullary terms.  For each $k \in \{1,\ldots,n\}$ splitting the
subterm $e_k$ yields $p_k$ and $l_k.r_k$ where:
\[
	\reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
\]
Correspondingly, splitting the $n$-ary $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where:
\[
	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
\]
Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
$l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.

Again, this holds even for nullary terms and under binders.
In particular, \texttt{case} splits according to the rule:
\[
\infersplittwo	{	\begin{array} {l}
						\caseof{e_1}{x_2.e_2}{x_3.e_3} \splittwosym \\
						\hspace{5em} \pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}		
					\end{array}}	
				{\sub 1 ? & \sub 2 ? & \sub 3 ?} 								
\]
Note how the precomputations are pulled out from underneath the binder
into a tuple that is always executed in the first stage. This occurs
regardless of what branch is eventually executed in the second stage.

\paragraph{Next and Prev.}
Since \texttt{next} and \texttt{prev} do nothing more than convert between 
second-stage terms at \bbtwo\ and second-stage terms encapsulated within $\fut$s at \bbonem,
their splitting rules are relatively simple.
Given an $\splittwo e ? p l r$ and $\diatwo e q$, the splitting invariant says,
\[
	\reduce p b \text{ and } [b/l]r \equiv q
\]
The splitting invariant for $\splitone {\next {e'}} ? {c'} {l'.r'}$,
where $\diaone {\next {e'}} {{\hat y} \mapsto q} {\next {\hat y}}$
and $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$
then requires
\[
	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
\]
which can be fulfilled using $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.
